{\large Master Thesis}\\{\normalsize on

### 2011-07-20 doc/report.tex
m

\begin{proof}
Whenever \proc{DirectComp} reads a character of the string $s$, we instead use the random access described in \pref{lem:lz-random-access} to read the character from the compressed string. The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$. In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase. The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
\end{proof}

@@

We can try to completely avoid Predecessor and Level Ancestor when initializing the pair $(x,y)$. We could try to store all $N$ pairs, but that would require $O(N)$ space and defeat the purpose of compression. We could also just walk from the end of the string to the desired position, but that would bring the average case query time up to $O(N)$. Similarly to how \proc{DirectCompLookup} works, we can store some of these positions. If we store $t$ positions equally spaced across the string $s$, we would get $O(n+t)$ space and $O(N/t+|\LCE(i,j)|)$ query time. To keep the compressed space, we can set $t=n$, which gives $O(N/n+|\LCE(i,j)|) = O(\sqrt{N}+|\LCE(i,j)|)$ query time.

@@

We can reverse the string in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively. To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.

@@

Fingerprinting does not compress well either. We could LZ-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.

Since we have not been able to find improvements to the $O(N)$ worst case query time in $O(n)$ space using existing algorithms, we have looked into new algorithms. The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree. The LCE value is the length of a substring, which occurs at two positions $i$ and $j$ in the input string. This substring spans a number of phrases, where each phrase is a path in the tree of compression elements. If we can make fast LCE queries on such a tree, we can make fast LCE queries on highly compressable strings.

@@

We have looked into the cache performance of compressed \proc{DirectComp}. If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case. When the string has a better LZ-compression ratio, the I/O pattern becomes much worse. If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be next to each other in memory. This can happen many times, which gives $O(N)$ I/O's in worst case.

>>>>

\begin{proof}
Whenever \proc{DirectComp} reads a character of the string $s$, we instead use the random access described in \pref{lem:lz-random-access} to read the character from the compressed string. The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$. In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase. The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
\end{proof}

@@

We can try to completely avoid Predecessor and Level Ancestor when initializing the pair $(x,y)$. We could try to store all $N$ pairs, but that would require $O(N)$ space and defeat the purpose of compression. We could also just walk from the end of the string to the desired position, but that would bring the average case query time up to $O(N)$. Similarly to how \proc{DirectCompLookup} works, we can store some of these positions. If we store $t$ positions equally spaced across the string $s$, we would get $O(n+t)$ space and $O(N/t+|\LCE(i,j)|)$ query time. To keep the space within $O(n)$, we can set $t=n$, which gives $O(N/n+|\LCE(i,j)|) = O(\sqrt{N}+|\LCE(i,j)|)$ query time.

@@

We can reverse the string in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively. To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n_2)$ space for its tree of compression elements.

@@

Fingerprinting does not compress well either. We could LZ-compress each level of fingerprints just like the original string is compressed, but as the fingerprinted substrings become longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.

Since we have not been able to find improvements to the $O(N)$ worst case query time in $O(n)$ space using existing algorithms, we have looked into new algorithms. The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree. The LCE value is the length of two equivalent substrings, which occur at positions $i$ and $j$ in the input string. These substrings span a number of phrases. If we can make fast LCE queries on a tree of compression elements, we can make fast LCE queries on highly compressable strings.

@@

We have looked into the cache performance of compressed \proc{DirectComp}. If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's. When the string has a better LZ-compression ratio, the I/O pattern becomes much worse. If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be in the same block in memory. This can happen many times, which gives $O(N)$ I/O's in worst case.
